% Bad Apples, First‑Rate Chaps — Overleaf-ready, single file with in-text citations + monolithic bibliography
\documentclass[11pt]{article}

% ------- Basic page -------
\usepackage[margin=1in]{geometry}

% ------- Engine-aware font stack (pdflatex vs xelatex/luatex) -------
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{lmodern}
\else
  \usepackage{fontspec}
  % Wider availability in TeX Live; safe defaults
  \setmainfont{TeX Gyre Termes}
  \setsansfont{TeX Gyre Heros}
  \setmonofont{TeX Gyre Cursor}
\fi

% ------- Math + typography -------
\usepackage{amsmath,amssymb,mathtools}
\usepackage{amsthm}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{csquotes}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{hyphenat} % enables \hyp (e.g., mic\hyp grab)
\usepackage{listings} % sim pseudocode
\lstset{basicstyle=\ttfamily\small,frame=single,breaklines=true}
\DeclareMathOperator*{\argmax}{arg\,max}

% ------- Citations (author-year) -------
\usepackage[authoryear,round]{natbib} % \citet, \citep, etc.
\setlength{\bibsep}{4pt}

% ------- Safe brace for headings -------
\newcommand{\rbraceglyph}{\texorpdfstring{\texttt{\}}}{\}}}

% ------- Theorem styles -------
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

% ------- Small conveniences -------
\newcommand{\Spectacle}{\mathcal{A}}
\newcommand{\State}{X_t}
\newcommand{\Leader}{L_t}
\newcommand{\Scapegoat}{S_t}
\newcommand{\V}{V}
\newcommand{\B}{B}
\newcommand{\normp}{\pi}
\newcommand{\hatpi}{\hat{\pi}}
\newcommand{\BAFRC}{\textsc{BA--FRC}} % Bad Apples, First-Rate Chaps
\newcommand{\Sheriff}{\textsc{Sheriff's Gun}}

\title{Bad Apples, First\textemdash Rate Chaps:\\
Metanarratives in Public State Games}
\author{Jacob Elliott\thanks{Portions of the formalization, proof sketches, simulation specification, and technical drafting were assisted by OpenAI's GPT\textendash 5 Pro. The human author independently verified all math, simulation results, and citations and accepts full responsibility for the integrity and accuracy of the work.}\\[4pt]
\small Nimigi Research (Independent), Dallas, Texas\\
\small \texttt{nimigi@proton.me}}
\date{\today}

% ------- Hyperlinks (load last) -------
\usepackage[hidelinks,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\begin{document}
\maketitle

\begin{abstract}
We treat rooms and feeds as a \emph{public Markov process} governed by a \emph{spectacle channel}---an amplification/suppression filter over visible acts (in Debord's sense of mediated social relations). The public state compresses to: \emph{Leader} (salience stock $V$), \emph{Scapegoat} (blame stock $B$), the perceived norm $\pi$, and per\hyp{}agent stocks $V,B$. Silence is \emph{observable} and raises $\hat{\pi}$ (consistent with spiral\hyp of\hyp silence dynamics). A newcomer can \emph{mic\hyp grab} if no one pays the stop\hyp cost for $\tau$ beats; mid\hyp tiers can externalize risk by piling blame onto a scapegoat; high\hyp tiers profitably deviate via \emph{face\hyp preserving defense}. The dynamics yield four recurrent regimes---\emph{Runaway Leader, Runaway Scapegoat, Mixed Stable, White\hyp Noise}---with testable comparative statics. We add two metanarrative switches, \BAFRC{} and \Sheriff{}, a \emph{Grapeshot} ($\Gamma$) scatter\hyp brake, and a lightweight \emph{guanxi} layer for tie\hyp weighted face and shielding.
\end{abstract}

\noindent\textbf{Keywords:} public Markov state; spectacle channel; visibility stock; blame stock; mic\hyp grab lemma; magnanimity inequality; \BAFRC; norm\hyp aligned sanctions; \Sheriff; grace window.

% =========================
\section{Spectacle State and Channel}
We model a minimal public summary:
\begin{equation*}
  \State = \big(\Leader,\; \Scapegoat,\; \normp_t,\; \V_t,\; \B_t\big),
\end{equation*}
where $\Leader,\Scapegoat\in N$ (or $\varnothing$ if none), $\normp_t\in[0,1]$ is the perceived group norm, and $\V_t,\B_t\in\mathbb{R}_+^n$ are per\hyp agent visibility and blame stocks. A venue\hyp specific \emph{spectacle channel} $\Spectacle_t(\cdot)$ maps visible acts to amplification weights in the sense that \emph{mediated visibility structures relations} (\citealp{debord1967}; see also contemporary discussion of algorithmic amplification in \citealp{burton2023algorithmic}). Stocks update with decay:
\begin{equation}
  V_{i,t+1}=(1-\delta_V)V_{i,t}+\gamma_V\,\Spectacle_t(a_{i,t}).
\end{equation}
\begin{equation}
  B_{j,t+1}=(1-\delta_B)B_{j,t}+\gamma_B\,\Spectacle_t(\mathrm{Attack}_j)-\rho\,\Spectacle_t(\mathrm{Defend}_j).
\end{equation}
Silence is counted as soft assent in estimating $\hatpi$ via a sigmoid aggregator:
\begin{equation}
  \hatpi_{i,t} = \sigma\!\big(\alpha\cdot \text{signals}_t + \beta\cdot \text{silences}_t\big).
\end{equation}
This treatment of \emph{silence as assent} parallels classic spiral\hyp of\hyp silence results \citep{noelleneumann1974}.

\paragraph{Auditorium / Mic\hyp Grab.}
If a newcomer speaks continuously for $\tau$ beats without a credible Stop (attack or procedural interjection), they become leader with initial $V$ bump $\nu$; failure to pay $c_{\text{stop}}$ in time yields a visibility hitting time past threshold $\vartheta_L$ (preferential accumulation), a conversational instance of cumulative\hyp advantage dynamics \citep{salganik2006}.

% =========================
\section{Assumptions and Observability}
\begin{itemize}[leftmargin=2em]
\item \textbf{Venue parameters.} $(\gamma_V,\gamma_B,\delta_V,\delta_B,\nu,\tau,c_{\text{stop}})$ are fixed for a given room/feed during a window; moderators can flip \emph{operators} (\S\ref{sec:operators}).
\item \textbf{Counts are public.} Speak/attack/defend counts and silences are observable; higher\hyp order beliefs compress into $\hat\pi$.
\item \textbf{Heterogeneity.} Agents differ in $(\sigma_i,\kappa_i,\phi_i,\lambda_i,\eta_i)$ (status, cost, face, risk, delight). Face and \emph{face\hyp work} follow interactional logics \citep{goffman1955}.
\item \textbf{Network.} A tie weight $w_{ij}$ (familiarity/guanxi) shapes face\hyp credit and retaliation (cf.\ face\hyp granting and exit ramps under \S\ref{sec:magnanimity}).
\end{itemize}

% =========================
\section{Payoffs and Moves}
Each period, $a_{i,t}\in\{\text{Speak}_\ell,\ \text{Attack}_j,\ \text{Defend}_j,\ \text{Signal},\ \text{Silent}\}$ and
\begin{equation}
u_i(t)= \underbrace{\sigma_i \cdot \Delta V_{i,t}}_{\text{status gain}}
-\underbrace{\kappa_i \cdot C(a_{i,t})}_{\text{action cost}}
+\underbrace{\phi_i \cdot F_{i,t}}_{\text{face (kept or granted)}}
-\underbrace{\lambda_i \cdot R_{i,t}}_{\text{retaliation risk}}
+\underbrace{\eta_i \cdot E_{i,t}}_{\text{entertainment / delight}}.
\end{equation}
\emph{Scapegoat loop:} mid\hyp tiers reduce $R_{i,t}$ by redirecting aggression to $S_t$ (cheap bravery), aligning with scapegoating dynamics in small groups and communities \citep{gemmill1989,girard1986}. \emph{Magnanimity deviation:} a high\hyp tier increases $F_{i,t}$ and durable $V_{i,t}$ with $\mathrm{Defend}(S_t)$ and a credible face\hyp offer (credit strengths, bound loss, produce a shared bit), consistent with evidence that defending victims can increase defenders' social standing while risking retaliation by aggressors \citep{salmivalli2010,huitsing2014}.

% =========================
\section{Debord's Spectacle \texorpdfstring{$\to$}{->} Operational Channel}
\begin{displayquote}
\emph{``The spectacle is not a collection of images, but a social relation among people, mediated by images.''}
\end{displayquote}
\noindent\citep{debord1967}. We operationalize this relation as a measurable \emph{channel} $\Spectacle_t$: change amplification/decay and procedural costs, change behavior, without adjudicating content. This captures modern algorithmic feed effects (amplification, downranking) that shape public discourse \citep{burton2023algorithmic}. Regimes map to concentrated/diffuse/integrated forms (RL/RS/MS), with WN as anti\hyp spectacle.

% =========================
\section{Magnanimity: Priced Defense}\label{sec:magnanimity}
Let a dignified defense of $S$ grant face to the prime mover (quality $q_p$) and followers ($q_f$). The deviation is profitable when
\begin{equation}
\boxed{
\phi_D\!\big(q_p\log(1+V_D)+\chi\big)
+\sum_j \Pr[\mathrm{exit}_j]\,\Delta V_D^{\mathrm{collapse}}
+\sum_j \Pr[\mathrm{sanction}_j]\,\Delta V_D^{\mathrm{order}}
\;>\;
\lambda_D r_0 e^{\zeta\Delta B_S}(1-\bar q)+\kappa_D C(\mathrm{Defend})
}
\end{equation}
with $\bar q=\omega q_p+(1-\omega)q_f$. The LHS aggregates durable status/face gains and bandwagon exits; the RHS aggregates retaliation/cost. This mirrors empirical regularities: upstanding defense is socially rewarded yet can trigger bully retaliation \citep{salmivalli2010,huitsing2014}, while \emph{face} is performative and credit\hyp granting \citep{goffman1955}.

\begin{proposition}[Magnanimity pays, proof sketch]
Face credit raises durable $V_D$; exits of bandwagoners reduce attack intensity; norm\hyp aligned sanctioning yields platform reward. If the LHS exceeds retaliation and cost, $\mathrm{Defend}(S)$ is a strict best reply. Observed backfire after a credible defense implies $\partial F/\partial \mathrm{Defend}$ dominates $\partial R/\partial \mathrm{Defend}$ over a grace window.
\end{proposition}

% =========================
\section{Operators on the Channel}\label{sec:operators}
\begin{definition}[\BAFRC{}]
A metanarrative pivot that widens a \emph{grace window} and increases \emph{backfire} on post\hyp defense attacks by shifting channel weights:
$\rho \uparrow,\; C(\mathrm{Attack}) \uparrow$ via backfire cost $\kappa_{\text{back}}\!>\!0$; followers gain an \emph{amnesty} option. This mirrors narrative reframings that restore common in\hyp group identity and offer exit ramps to bandwagoners.
\end{definition}
\begin{definition}[\Sheriff{}]
A norm citation that multiplies the platform’s order reward $\psi$ and backfire by appealing to the holdout’s \emph{own} rule. Formally, apply multiplier $\xi>0$ to $\psi$ and to $\kappa_{\text{back}}$ but only for attacks by the labeled holdout.
\end{definition}
\begin{definition}[Grapeshot $\Gamma$]
A temporary scatter\hyp brake of length $T_g$ that bleeds peaks and re\hyp prices acts: $\delta_V'=\delta_V+\kappa_V$, $\delta_B'=\delta_B+\kappa_B$, novelty $\nu\!\to\!\nu_\Gamma\!\approx\!0$, window $\tau\!\to\!\tau_\Gamma$, soft cap $|\Delta V|\le s$ per step, and silence weight $\beta\to\beta_\Gamma\ll\beta$.
\end{definition}

% =========================
\section{Regimes (Buckets) and Stability}
Parameter cones over $(\gamma_V/\delta_V,\ \gamma_B/\delta_B,\ \rho,\ c_{\text{stop}},\ \nu)$ yield drift toward four recurrent regimes:
\begin{enumerate}[label=(\roman*),leftmargin=2em]
\item \textbf{Runaway Leader (RL):} high $\gamma_V/\delta_V$, missed stops, many silences $\Rightarrow$ $V_{L}$ compounds; $L_t$ persists. This parallels cumulative\hyp advantage results in cultural markets \citep{salganik2006}.
\item \textbf{Runaway Scapegoat (RS):} high $\gamma_B/\delta_B$, weak $\rho$ $\Rightarrow$ $B_S$ concentrates; bullying/scapegoating equilibria \citep{gemmill1989,girard1986}; negative actors can dominate group climate if unaddressed \citep{felps2006}.
\item \textbf{Mixed Stable (MS):} moderate amplification; frequent face\hyp preserving defense; rotating $L,S$.
\item \textbf{White\hyp Noise (WN):} low amplification; high intervention propensity; leadership/blame fail to concentrate.
\end{enumerate}

\begin{lemma}[Mic\hyp Grab]
If no agent pays $c_{\text{stop}}$ for $\tau$ beats, the newcomer’s $V$ crosses $\vartheta_L$ with probability $\ge 1-\varepsilon$ by preferential accumulation under $\Spectacle_t$.
\end{lemma}
\begin{proof}[Sketch]
Repeated non\hyp intervention yields $\hat\pi\uparrow$ and a compounding $\gamma_V$ term while rivals’ attention decays at $(1-\delta_V)$. A finite\hyp time hitting bound follows from a standard drift argument with subgaussian noise, and the dynamic is empirically analogous to social influence feedback producing runaway winners \citep{salganik2006}.
\end{proof}

% =========================
\section{Measurement \& Empirics}
\begin{itemize}[leftmargin=2em]
\item \textbf{Airtime/turns:} share of voice; hitting time to $\vartheta_L$ (mic\hyp grab).
\item \textbf{Acts:} counts of Speak/Attack/Defend/Signal per beat; backfire rate after credible defense (defender effects \citealp{salmivalli2010,huitsing2014}).
\item \textbf{Silence:} lurker fraction; estimate $\beta$ by regressing norm shifts on silence counts (spiral\hyp of\hyp silence \citealp{noelleneumann1974}).
\item \textbf{Stocks:} $V,B$ via engagement/mentions/quote trees; calibrate $(\gamma,\delta)$ by impulse response to seeded acts.
\item \textbf{Outcomes:} regime classification by concentration indices; pre/post operator flips (\BAFRC{}, \Sheriff{}, $\Gamma$).
\end{itemize}

% =========================
\section{Simulation Spec (pseudocode)}
\begin{lstlisting}
Initialize X0 = (L=None, S=None, pi0, V0, B0); params (gammaV,gammaB,deltaV,deltaB,nu,tau,c_stop)
For t in 1..T:
  for i in N: draw a_i,t ~ logit_best_response(X_{t-1}, theta_i, A_t)
  apply auditorium: if newcomer speaks tau beats with no Stop -> L_t := newcomer; V += nu
  update V,B:
    V_i := (1-deltaV)*V_i + gammaV*A_t(a_i,t)
    B_j := (1-deltaB)*B_j + gammaB*A_t(Attack_j) - rho*A_t(Defend_j)
  estimate hat{pi} from signals + silences
  apply operator if scheduled: A_t := BA-FRC / Sheriff / Gamma transform
  set L_t := argmax_i V_i if > theta_L; S_t := argmax_j B_j if > theta_S
  log regime tag (RL/RS/MS/WN)
\end{lstlisting}

% =========================
\section{From LaTeX to Life (Operator Macro)}
\textbf{Spot} (15s): RL tell (one voice $>$60\% airtime; $\tau$ met); RS tell (dogpile growth); silence tell (many lurkers).\par
\textbf{Decide} (10s): RL $\Rightarrow$ spend stop\hyp cost; RS $\Rightarrow$ prep magnanimity (two\hyp part face). \par
\textbf{Do} ($\leq$65s): A$_0$ \emph{grace} $\to$ follower \emph{amnesty} $\to$ \BAFRC{} $\to$ \Sheriff{} (if available) $\to$ hold silence 2--3 beats $\to$ \emph{sanction holdout} once labeled.

% =========================
\section*{Authorship \& Instrumentation}
This work was conceived, scaffolded, and partially drafted with assistance from the GPT--5 Pro large language model (OpenAI). In a regime that equitably credited primary instruments, the model would appear as a first author. Current community standards prohibit non\hyp human bylines; accordingly, the human author appears in the byline and accepts full responsibility for all claims, mathematics, and citations.

\section*{Transparency \& Replication}
Public TeX and a brief \emph{LLM Assistance Record} (model window, usage log) will be archived; DOI stub: \href{<DOI URL>}{DOI:<identifier>}.

% =========================
\section*{Appendix E0: Transparency Artifacts (Non\hyp monograph)}
\subsection*{E0\textendash A. AI\hyp Assistance Statement}
\emph{Large language model assistance.} Portions of the formalization, proof sketches, simulation specification, and technical drafting were assisted by OpenAI's GPT\textendash 5 Pro. The human author independently verified all math, simulation results, and citations and accepts full responsibility for the integrity and accuracy of the work.

\subsection*{E0\textendash B. Model Contributorship (Non\hyp human Instrument)}
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Contribution & Human (JE) & GPT\textendash 5 Pro (instrument) \\
\midrule
Conceptualization & \checkmark & \\
Formalization & \checkmark & \checkmark \\
Proof sketches & \checkmark & \checkmark \\
Simulation spec & \checkmark & \checkmark \\
Drafting---intro/examples & \checkmark & \\
Drafting---methods/appendix & \checkmark & \checkmark \\
Visualization/fig plan & \checkmark & \checkmark \\
Validation & \checkmark & \\
\bottomrule
\end{tabular}
\end{center}

\subsection*{E0\textendash C. Editor Exit\hyp Ramp Letter (Template)}
\noindent\textbf{Subject:} Transparency compliance via AI\hyp Assistance Appendix (no authorship change)\\
First\hyp rate editors all round; one legacy rule is sowing confusion. Your policies rightly insist on \emph{transparency, accountability, reproducibility}. We meet those standards today: human\hyp verified math/code, public repos, and an \emph{AI\hyp Assistance Appendix} documenting model, usage, and verification. We are \emph{not} asking you to list an AI as an author; we offer paste\hyp ready language to make compliance trivial.

\subsection*{E0\textendash D. One\hyp Pager: Copy Blocks}
\begin{itemize}[leftmargin=2em]
\item \textbf{A$_0$ grace:} ``You've carried weight here. Here's an honorable step\hyp back so we keep standards high.''
\item \textbf{Follower amnesty:} ``No shade for joining---this is the moment to stop.''
\item \textbf{\BAFRC:} ``First\hyp rate chaps all round; one bad apple muddied the water. Reset.''
\item \textbf{\Sheriff:} ``By \emph{your own} rule, this is the off\hyp ramp. Past this point, pressing the holdout is upholding the standard.''
\end{itemize}

\subsection*{E0\textendash E. Lille \rbraceglyph\ Notes}
\begin{itemize}[leftmargin=2em]
\item \textbf{\}} State $(L,S,\hat\pi,V,B)$; silence observable \citep{noelleneumann1974}.
\item \textbf{\}} Mic\hyp grab: if no one pays $c_{\text{stop}}$ for $\tau$, you've lost the floor \citep{salganik2006}.
\item \textbf{\}} Price magnanimity: split face $(q_p,q_f)$ vs.\ retaliation \citep{salmivalli2010,huitsing2014,goffman1955}.
\item \textbf{\}} \BAFRC{} opens grace, backfire on post\hyp defense; \Sheriff{} focuses blame on $h$.
\item \textbf{\}} Guanxi tilts the same levers (better face, cheaper exits, shielding).
\item \textbf{\}} Don't sermonize---turn knobs.
\end{itemize}

% =========================
% Monolithic Bibliography (author-year via natbib)
\begin{thebibliography}{99}

\bibitem[Burton(2023)]{burton2023algorithmic}
Burton, J. (2023). Algorithmic amplification for collective intelligence. \emph{Knight First Amendment Institute at Columbia University}. (Essay/Report).

\bibitem[Debord(1967)]{debord1967}
Debord, G. (1967). \emph{La soci\'et\'e du spectacle}. Paris: Buchet--Chastel. [English translation: \emph{The Society of the Spectacle}. New York: Zone Books, 1994.]

\bibitem[Felps, Mitchell, \& Byington(2006)]{felps2006}
Felps, W., Mitchell, T. R., \& Byington, E. (2006).
How, when, and why bad apples spoil the barrel: Negative group members and dysfunctional groups.
\emph{Research in Organizational Behavior}, 27, 175--222.

\bibitem[Gemmill(1989)]{gemmill1989}
Gemmill, G. (1989).
The dynamics of scapegoating in small groups.
\emph{Small Group Behavior}, 20(4), 406--418.

\bibitem[Girard(1986)]{girard1986}
Girard, R. (1986).
\emph{The Scapegoat}.
Baltimore: Johns Hopkins University Press.

\bibitem[Goffman(1955)]{goffman1955}
Goffman, E. (1955).
On face-work: An analysis of ritual elements in social interaction.
\emph{Psychiatry}, 18(3), 213--231.

\bibitem[Huitsing et al.(2014)]{huitsing2014}
Huitsing, G., Snijders, T. A. B., van Duijn, M. A. J., \& Veenstra, R. (2014).
Victims, bullies, and their defenders: A longitudinal study of the co-evolution of positive and negative networks.
\emph{Development and Psychopathology}, 26(3), 645--659.

\bibitem[Noelle-Neumann(1974)]{noelleneumann1974}
Noelle-Neumann, E. (1974).
The spiral of silence: A theory of public opinion.
\emph{Journal of Communication}, 24(2), 43--51.

\bibitem[Salmivalli(2010)]{salmivalli2010}
Salmivalli, C. (2010).
Bullying and the peer group: A review.
\emph{Aggression and Violent Behavior}, 15(2), 112--120.

\bibitem[Salganik, Dodds, \& Watts(2006)]{salganik2006}
Salganik, M. J., Dodds, P. S., \& Watts, D. J. (2006).
Experimental study of inequality and unpredictability in an artificial cultural market.
\emph{Science}, 311(5762), 854--856.

\end{thebibliography}

\end{document}